Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-01-30 paper-1/manuscript.tex
Editing

\abstract{
    Sensory areas of cortex respond more strongly to infrequent stimuli when these violate previously established regularities, a phenomenon known as deviance detection (DD). Previous modeling work has mainly attempted to explain DD on the basis of synaptic plasticity and tightly constrained network structure. However, a large fraction of cortical neurons also exhibit intrinsic plasticity, an underexplored potential mechanism, and recent results show DD in dissociated neuronal culture, raising doubts about the necessity of network structure.
    Here, we investigate DD in a spiking neuronal model with spatially constrained connectivity and two types of short-term plasticity, fast synaptic short-term depression (STD) and slower threshold adaptation (TA). We probe the model with oddball sequences at an intermediate inter-stimulus interval and assess DD.
    We find that TA alone is sufficient to generate DD signals, and identify an interplay between suppression and its recovery as the underlying mechanism.
    Further, we find that the full model with both STD and TA yields a DD signal that exceeds the sum of the DD signals in ablated models with either STD or TA turned off. We identify the cause of this in an interaction where the fast-decaying STD increases local suppression near the frequently stimulated neurons, reducing the stimulus response, and thereby indirectly promotes global recovery of TA.
    We suggest that similar interactions between various forms of plasticity are likely to shape how sensory networks process regularities.
}

@@

Deviance detection (DD) is the ability of neural systems to identify statistically surprising sensory inputs or violations of established regularities and to respond to this surprise with an increased signal. Deviance detection is a critical component of any signal processing cascade, and is closely related to the concept of error signals in the predictive processing framework \citep{Bendixen2012-nx, Khouri2015-gr, Fong2020-em}. DD has been studied particularly in auditory processing with approaches from single-unit recordings \citep{Taaseh2011-gg} to EEG {Winkler2012-pr}. Impairments in DD have been reported in various disorders \citep{Fong2020-em}, including autism \citep{Schwartz2018-bc} and schizophrenia \citep{Tada2019-lj}, highlighting the need to better understand its mechanistic underpinnings.

>>>>

\abstract{
    Sensory areas of cortex respond more strongly to infrequent stimuli when these violate previously established regularities, a phenomenon known as deviance detection (DD). Previous modeling work has mainly attempted to explain DD on the basis of synaptic plasticity and tightly constrained network structure. However, a large fraction of cortical neurons also exhibit intrinsic plasticity, an underexplored potential mechanism, and recent results show DD in dissociated neuronal culture, raising doubts about the necessity of network structure.
    Here, we investigate DD in a spiking neuronal model with spatially constrained connectivity and two types of short-term plasticity, fast synaptic short-term depression (STD) and slower threshold adaptation (TA). We probe the model with oddball sequences at an intermediate inter-stimulus interval and assess DD.
    We find that TA alone is sufficient to generate DD signals and identify an interplay between suppression and its recovery as the underlying mechanism.
    Further, we find that the full model with both STD and TA yields a DD signal that exceeds the sum of the DD signals in ablated models with either STD or TA turned off. We identify the cause of this in an interaction where the fast-decaying STD increases local suppression near the frequently stimulated neurons, reducing the stimulus response, and thereby indirectly promotes global recovery of TA.
    We suggest that similar interactions between various forms of plasticity are likely to shape how sensory networks process regularities.
}

@@

Deviance detection (DD) is the ability of neural systems to identify statistically surprising sensory inputs or violations of established regularities and to respond to this surprise with an increased signal. Deviance detection is a critical component of any signal processing cascade and is closely related to the concept of error signals in the predictive processing framework \citep{Bendixen2012-nx, Khouri2015-gr, Fong2020-em}. DD has been studied particularly in auditory processing with approaches from single-unit recordings \citep{Taaseh2011-gg} to EEG {Winkler2012-pr}. Impairments in DD have been reported in various disorders \citep{Fong2020-em}, including autism \citep{Schwartz2018-bc} and schizophrenia \citep{Tada2019-lj}, highlighting the need to better understand its mechanistic underpinnings.
